Basics of Hardware Cache Optimization

### 2017-05-24 tex/types.tex
Expand content and tweak stuff

My laptop's AMD E-450 CPU has cores with \alts{an \gls{l1d} cache of 32 KiB, 32 KiB of
\gls{l1d} cache} and a unified \gls{l2} cache of 512 KiB each.%
\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
We can both verify these sizes and get \alts{a reasonably good measure, a rough measure,
an approximation} of the access times by profiling
\cref{lst:access-times}
% the \alts{program, listing} below
for different values of \mintinline{text}{SIZE}.%
% FIXME: the label is wrong: should be "Appendix" but is "Section".
\footnote{\Cref{app:cycles} details how.}
% It reads random memory addresses
% The program randomly reads
% locations
% The program repeatedly \alts{reads, accesses} random elements
This program repeatedly \alts{reads, accesses} elements
from \alts[2]{an array of the configured size, a thusly sized array}
in random \alts{order, sequence, succession}.
To do this
\alts{with minimal overhead, efficiently}, the array \alts{is first set up, acts} as a
circular, singly linked list where every element except the last \alts{points to a, has a}
random successor.  When compiled with \mintinline{text}{-DBASELINE}, only this
initialization is done.
\Cref{fig:access-times} shows the deltas of CPU cycles used when and when not having
defined \mintinline{text}{BASELINE}.
I divided by \mintinline{text}{N} to get the cycles spend on each loop iteration.

@@

% Actually, don't use a float.  The listing should be allowed to span multiple pages which
% floats aren't.
%
% [1]: https://tex.stackexchange.com/q/14522/#comment484569_75880
% [2]: https://tex.stackexchange.com/q/175650
%      "How to allow page break inside a float environment?"
% [3]: https://tex.stackexchange.com/q/12428
\inputminted[firstline=9]{c}{access-times/access-times.c}
\captionof{listing}{}
% \captionof{listing}{Random Reads\label{lst:access-times}}
\label{lst:access-times}

@@

>>>>

My laptop's AMD E-450 CPU has cores with \alts{an \gls{l1d} cache of 32 KiB, 32 KiB of
\gls{l1d} cache} and a unified \gls{l2} cache of 512 KiB each.%
\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
We can both verify these sizes and get \alts{a reasonably good measure, a rough measure,
an approximation} of the access times by profiling
\cref{lst:access-times}
% the \alts{program, listing} below
for different values of \mintinline{text}{SIZE}.%
% FIXME: the label is wrong: should be "Appendix" but is "Section".
\footnote{\Cref{app:cycles} details how.}
% It reads random memory addresses
% The program randomly reads
% locations
% The program repeatedly \alts{reads, accesses} random elements
This program repeatedly \alts{reads, accesses} elements
from \alts[2]{an array of the configured size, a thusly sized array}
in random \alts{order, sequence, succession}.
To do this
\alts{with minimal overhead, efficiently}, the array \alts{is first set up, acts} as a
circular, singly linked list where every element except the last \alts{points to a, has a}
random successor.  When compiled with \mintinline{text}{-DBASELINE}, only this
initialization is done.

@@

% XXX: hacks!  FIXME: I want to have the same amount of space after adding a caption with
% \captionof to a minted listing as there would be after a normal floating minted listing.
% See <https://tex.stackexchange.com/a/162074>.
% \captionsetup[listing]{aboveskip=5pt, belowskip=\baselineskip}

% Actually, don't use a float.  The listing should be allowed to span multiple pages which
% floats aren't.
%
% [1]: https://tex.stackexchange.com/q/14522/#comment484569_75880
% [2]: https://tex.stackexchange.com/q/175650
%      "How to allow page break inside a float environment?"
% [3]: https://tex.stackexchange.com/q/12428
\begin{center} % XXX: hack to get normal spacing after the caption and before the listing.
   \inputminted[firstline=12]{c}{access-times/access-times.c}
   \captionof{listing}{}
   % \captionof{listing}{Random Reads\label{lst:access-times}}
   \label{lst:access-times}
\end{center}

@@

\Cref{fig:access-times} shows the \alts{difference, deltas} of CPU cycles used when and
when not having defined \mintinline{text}{BASELINE}.  That is, only the cycles used by the
main loop are counted, not those for initialization.
I divided by \mintinline{text}{N} to get the cycles spend \alts{per, on each} loop
iteration.

Up to 32 KiB, each access takes almost exactly 3 cycles.%
\footnote{TODO: table of actual data in appendix?}
This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
the time increases to about 3.4 cycles.  This is not surprising since the cache is shared
with other processes and the operating system, so some of our data gets evicted.  The
first dramatic increase happens at 64 KiB followed by smaller increases at 128 and 256
KiB.  I suspect we are seeing a mixture of \gls{l2} and \gls{l1d} accesses, with less and
less \gls{l1d} hits and an \gls{l2} access time of around 25 cycles.

The values from 512 KiB to 128 MiB \alts{exhibit, follow} a similar pattern.  The relative
increase when the array size matches that of the \gls{l2} is more striking than for the
\gls{l1d} before; possibly because \gls{l2} is a unified cache that also holds
instructions.  Eventually, more and more accesses go to main memory, \alts{causing,
incurring} delays of up to 200 cycles.

TODO~\cite[cf.][17]{drepper2007}.
